---
title: "Stat330 Project"
format: html
editor: visual
---

For this project, our group selected the Portugese Student Performance dataset from the University of California Irvine. This dataset contains 33 variables. These range from the occupations of each parent, the education level of each parent, the free time each student has, income, age, address, and even whether the student is in a romantic relationship. With so many variables, it seems appropriate to thin the data set to a number of variables.

For this analysis, we are considering the following variables:

| Variable Name | Type        | Description                                                                     |
|-----------------------------|------------------|-------------------------|
| Sex           | Categorical | Student's Sex                                                                   |
| Age           | Continuous  | Student's Age in Years (Rounded Down)                                           |
| Address       | Categorical | Whether a Student lives in a Rural Area or Urban Area                           |
| Famsize       | Categorical | Whether a Student's Family is Less than or Equal to Three or Greater than Three |
| Pstatus       | Categorical | Parent's Cohabitation Status                                                    |
| Failures      | Continuous  | Number of Past Class Failures                                                   |
| Schoolsup     | Categorical | Extra Educational Support                                                       |
| Famsup        | Categorical | Family Educational Support                                                      |
| Higher        | Categorical | Aspires to Higher Education                                                     |
| Internet      | Categorical | Has Internet Access at Home                                                     |
| Romantic      | Categorical | Has a Romantic Relationship                                                     |
| Absences      | Continuous  | Number of School Absences                                                       |
| G1            | Continuous  | First Period Grade                                                              |
| G2            | Continuous  | Second Period Grade                                                             |
| G3            | Continuous  | Final Grade                                                                     |

```{r}
#This code does not need to be run again
#This code can almost certainly be removed from the report
data <- read.csv('student-mat.csv', sep = ';')
dataCleaned <- data[-c(1, 7:14, 18:20, 24:29)]

dataCleaned$female = dataCleaned$sex == 'F'
dataCleaned$rural = dataCleaned$address == 'R'
dataCleaned$bigFamily = dataCleaned$famsize == 'GT3'
dataCleaned$parentsApart = dataCleaned$Pstatus == 'A'
dataCleaned$extraSchool = dataCleaned$schoolsup == 'yes'
dataCleaned$familyHelp = dataCleaned$famsup == 'yes'
dataCleaned$seeksCollege = dataCleaned$higher == 'yes'
dataCleaned$internet = dataCleaned$internet == 'yes'
dataCleaned$significantOther = dataCleaned$romantic == 'yes'

dataCleaned <- dataCleaned[-c(1, 3:5, 7:11)]

write.csv(dataCleaned, "cleanedMathData.csv", row.names = FALSE)
```

```{r}
fullDataSet <- read.csv('cleanedMathData.csv')
```

